---
title: "Comparison of splicing events of MAJIQ with rMATS"
author: "Maria Kondili"
date: "2026-05-27"
header-includes:
  - \usepackage{fvextra} 
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines=TRUE,commandchars=\\\{\}}
output: html_document
editor_options: 
  chunk_output_type: console
---


```{r clean_env }

rm(list = ls())  

```

```{r chunk_options, include=FALSE}

knitr::opts_chunk$set(echo = FALSE,       # FALSE: no code will be shown in html/pdf report
                      eval = TRUE,
                      fig.dim = c(10,7),  # for larger space for a graph
                      pillar.sigfig = 5,  # number of decimals to show from a tibble
                      tidy.opts=list(width.cutoff=70),
                      tidy=TRUE) #wrap long-messages in two-lines
                    
```


```{r libs}
library(tidyverse)
library(dplyr)
library(purrr)
```


```{r dirs}
epissav_dir <- "~/MYO_WORKSPACE/PROJECTS/EPIS_SAV-JohnRendu/"
majiq_wsp <- paste0(epissav_dir,"MAJIQ/")
work_dir <- paste0(majiq_wsp,"Scripts/R_dowstream/")
#setwd(work_dir)

```

#### MAJIQ-results (merged_tables)

```{r read_majiq_res}

modulizer_dir <- paste0(majiq_wsp ,"majiq_wsp/modulizer/changing_dpsi_0.3/")

majiq_mrg_dir <- paste0(modulizer_dir,"Merged/")

samples <- list.files(modulizer_dir, pattern="C00*")

majiq_mrg_filenames <- list.files(majiq_mrg_dir,full.names = T)

majiq_mrg_tables <-  map(majiq_mrg_filenames, ~read_delim(.x, delim="\t",
                                                          show_col_types = F,
                                                          col_names = TRUE)
                         )

## use the filenames as sample-names for the list of tables 
names(majiq_mrg_tables) <- str_remove(list.files(majiq_mrg_dir),".tsv")


mjq.num_rows_before_filt <- tibble( sample = names(majiq_mrg_tables), 
                                    num_rows = unlist(map(majiq_mrg_tables, ~nrow(.x))) )

cat("Nb of rows per sample, before filtering:")
mjq.num_rows_before_filt 

```

> Filter Majiq-results, |gr2-gr1_median_deltapsi| > 0.3 

```{r filter_majiq_mrg_tables }

majiq_dpsi_co <- 0.3

majiq_mrg_tables.filt <-  map(majiq_mrg_filenames, ~read_delim(.x, delim="\t",
                                                          show_col_types = F,
                                                          col_names = TRUE) %>% 
                                dplyr::filter( abs(`grp2-grp1_median_dpsi`) > majiq_dpsi_co )
                         )

## give names to list:
names(majiq_mrg_tables.filt) <- names(majiq_mrg_tables)

mjq.num_rows_after_filt <- tibble( sample = names(majiq_mrg_tables), 
                                  num_rows = unlist(map(majiq_mrg_tables.filt, ~nrow(.x)) ) )

cat("Nb of rows per sample, after filtering:")
mjq.num_rows_after_filt

```

#### rMATS-results (merged-tables)

```{r rmats_dir}

rmats_dir <- paste0(epissav_dir,"rMATS/Downstream_Analysis/Outputs/one_group_stat_off/Downstream_Outputs/final_rmats_merged/only_merged_tables/")

```

> Read symlinks of files from rmats

```{r resolve_symlinks_rmats}

## Resolve the shortcut: 
#--- FUNCTION-----#
resolve_lnk <- function(filepath) {
            system2("powershell", 
                    args = c("-command",sprintf(
                    "(New-Object -ComObject WScript.Shell).CreateShortcut('%s').TargetPath",
                    normalizePath(filepath) )),
            stdout = TRUE )
}

## to be able to read it :
# target <- resolve_link(symlink_file)
# df <- read_delim(target, delim = "\t")


```


```{r rmats_tables }

## Capture real files from symlink
shortcuts <- list.files(rmats_dir, pattern="- Shortcut.lnk", full.names = TRUE)

rmats_mrg_filenames <- map(shortcuts, ~resolve_lnk(.x) )

rmats_mrg_tables   <-  map(rmats_mrg_filenames, ~read_delim(.x, delim="\t",
                                                          show_col_types = F,
                                                          col_names = TRUE)
                         )

mrg_filenames <- unlist(map(rmats_mrg_filenames, ~basename(path = .x[[1]]) ))


## Give names to list: from "merged_modif_ASE_sample44_FDRsorted.tsv"--> extract "sample44" 
names(rmats_mrg_tables) <- c(substr(mrg_filenames[1:5], start=18,stop=25),
                             substr(mrg_filenames[6:17],start=18,stop=24) )

```

> Rename rmats-samples to match MAJIQ-sample_names

```{r}

names(rmats_mrg_tables) <- c("C002I44","C002I45","C002I47","C002I4F", "C002I4G",
                             "C002I3K", "C002I3L","C002I3M","C002I3N","C002I3P","C002I3Q",
                             "C002I3R", "C002I3S","C002I3U","C002I3W","C002I3Y","C002I3Z")

samples.rmats <- names(rmats_mrg_tables) 
```

> Filter rMATS : |dPSI| > 0.5 & FDR <= 0.05

```{r filter_rmats}

rmats_dpsi_co <-  0.5


## Rows before: 
num_rows_before_filt <- tibble( sample = samples.rmats, 
                                num_rows= unlist(map(rmats_mrg_tables, ~nrow(.x))) )


### Filter by dPSI :
rmats_mrg_tables.filt   <-  map(rmats_mrg_filenames, ~read_delim(.x, delim="\t",
                                                            show_col_types = F,
                                                            col_names = TRUE)  %>% 
                                  dplyr::filter(abs(dPSI) > rmats_dpsi_co & FDR <= 0.05 )
)

names(rmats_mrg_tables.filt) <- samples.rmats

## Rows after: 
num_rows_after_filt <- tibble( sample = samples.rmats, 
                               num_rows = unlist(map(rmats_mrg_tables.filt, ~nrow(.x))) )


```

#### Compare Events rmats/majiq

> Find common genes 

```{r names_common_genes }

find_common_genes <- function(rmats_tab, majiq_tab,sample_name) {

    #' Read rmats_tab[[sample]]$geneSymbol &
    #' majiq_tab[[sample]]$gene_name
    #' and intersect 
    cat("Sample:")
    print(sample_name)
    
    ##> TABLE_1
    rmats_genes <- rmats_tab[[sample_name]] %>%
                    pull(geneSymbol) %>%
                    unique() %>%
                    na.omit()
    
    print(rmats_genes[1:5])
    
    ## TABLE_2
    majiq_genes <- majiq_tab[[sample_name]] %>%
                    pull(gene_name) %>%
                    unique() %>%
                    na.omit()
    
    print(majiq_genes[1:5])
    
    common_genes <- intersect(rmats_genes, majiq_genes)
    return(common_genes)
}

# Find common genes for each sample
common_genes <- map(names(majiq_mrg_tables), 
                    ~find_common_genes( rmats_mrg_tables.filt,
                                        majiq_mrg_tables.filt,
                                        .x))

common_genes <- common_genes %>% set_names(names(majiq_mrg_tables))

```

> Save common-genes in file

```{r common_genes_in_file}

common_genes_outfile = glue::glue("Comparison_with_rMATS/Common_genes_rmats(dpsi={rmats_dpsi_co})-vs-majiq(dpsi={majiq_dpsi_co}).txt")

sink(common_genes_outfile)
  iwalk(common_genes, ~{
                        cat("###", .y, "\n") # sample
                        cat(.x, sep = "\n") # gene-names
                        cat("\n")
  })
sink()


if (file.exists(common_genes_outfile)) {
  print( glue::glue("File saved in file: {common_genes_outfile}"))
}

```


```{r test_common_genes_function,echo=F,eval=F }
## TEST:

for (s in samples.rmats) {
  
   majiq_genes <- majiq_mrg_tables.filt[[s]] %>%
        pull(gene_name) %>%
        unique() %>%
        na.omit()
   
  print(majiq_genes)
  
  rmats_genes <- rmats_mrg_tables[[s]] %>%
        pull(geneSymbol) %>%
        unique() %>%
        na.omit()
  
  print(rmats_genes)
}

```


```{r count_common_genes}


common_genes_counts <- tibble(sample = names(majiq_mrg_tables),
                              Nb_common = map_int(sample, ~length(common_genes[[.x]]) )
                              )

```

```{r write_down_results }

outdir <- "Comparison_with_rMATS/"
dir.create(outdir, showWarnings = FALSE)

readr::write_delim(common_genes_counts, 
                   file = paste0(outdir, "Table_common_genes_per_sample_after_filt.tsv"),
                   delim = "\t",append = FALSE)

```

#### Create VENN for showing common genes 

```{r plot_venn}

## developed with help by Claude-AI

library(ggVennDiagram)
library(purrr)
library(dplyr)


plot_venn_sample <- function( rmats_tab,majiq_tab, sample_name) {
  
  gene_list <- list(
                rMATS = rmats_tab[[sample_name]] %>%
                          pull(geneSymbol) %>% 
                          unique() %>% 
                          na.omit() %>% 
                          as.character(),
                MAJIQ = majiq_tab[[sample_name]] %>%
                          pull(gene_name) %>% 
                          unique() %>% 
                          na.omit() %>% 
                          as.character()
  )
  
  ggVennDiagram(gene_list,
                label        = "both",   # "both":shows count + percent
                label_alpha  = 0,
                label_size   = 5, #<-- percentage/count inside circle
                set_size     = 4, #<-- group(set)-label
                edge_size    = 0.5) +
    scale_fill_gradient(low = "#FFE4C4", high =  "#98F5FF") + 
    labs(title = paste0("Spliced Genes detected — ", sample_name)) +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, size = 13),
          plot.margin= margin(1,1,1,1))
    
   
}


venn_plots <- map(names(majiq_mrg_tables.filt), 
                  ~plot_venn_sample(rmats_mrg_tables.filt,
                                    majiq_mrg_tables.filt,
                                    .x))

# wrap_plots(venn_plots, ncol = 3) +
#   plot_annotation(
#     title = "rMATS vs MAJIQ — Common Spliced genes per sample",
#     theme = theme(plot.title = element_text(hjust = 0.5, 
#                   size = 12, face = "bold"))
#    )                  
#     

##----- Save with walk2 (multiple pngs) ------: 
# walk2(
#     venn_plots,
#     names(venn_plots),
#     ~ggsave(
#         filename = paste0(.y, "_venn_genes.png"),
#         plot = .x,
#         width = 6,
#         height = 5,
#         dpi = 300
#     )
# )


##-----show in Console ------- 
for (p in venn_plots){
      print(p)
}

##----- save in pdf ----------:
pdf(glue::glue("Comparison_with_rMATS/ggVennDiagrams_rMATS-filt_dpsi={rmats_dpsi_co}-vs-MAJIQ-filt_dpsi={majiq_dpsi_co}_onGenes_per_sample.pdf"),
    onefile = TRUE)
    for (p in venn_plots){
      print(p)
    }
dev.off()


```

####  Overlap MAJIQ-results (dpsi-filt) with  Muscle-Gene-Tables( Bonne,Rivier )

```{r overlap_with_muscle-genes-table}

## Read table with Genes per Disease: 
gene_tables_dir <- paste0(epissav_dir,"GeneTables_Bonne-Rivier/")

mrg_genetables_collapsed <- readr::read_delim(file =
                                          paste0(gene_tables_dir,
                                          "Merged_GeneTables_collapsed_diseases_03-06-26.txt"),
                                          delim = "\t",
                                          col_names = TRUE,
                                          na = "NA")

## Filter by GeneTable : 
majiq_tabs.filt_muscle_genes <- map(majiq_mrg_tables.filt, 
                                    ~(.x %>%  dplyr::filter(gene_name %in%
                                                           mrg_genetables_collapsed$GeneSymbol)))

##> Show genes per majiq_table: 
majiq_muscle_genes_per_sample <- map(majiq_tabs.filt_muscle_genes, 
                                     ~(unique(.x[["gene_name"]])) 
                                     )

## concatenate gene-names:
muscle_gene_names_in_majiq <- map(majiq_muscle_genes_per_sample, 
                                  ~ paste(.x,collapse=", ")  
                                  )
samples_of_muscle_gene_names <- names(muscle_gene_names_in_majiq)
muscle_gene_names_in_majiq <- unlist(muscle_gene_names_in_majiq)


muscle_gene_names_in_majiq.df <- tibble( "Sample" = samples_of_muscle_gene_names,
                                         "Gene_names" = muscle_gene_names_in_majiq)


cat("Muscle-Disease-Genes found in Majiq-spliced results,after dpsi-filtering :\n")
print(muscle_gene_names_in_majiq.df)

write_delim(muscle_gene_names_in_majiq.df,
            file="Muscle_Genes_in_Majiq_Results_per_sample.txt",
            delim="\t",
            col_names=T,
            na="NA")

```

#### MAJIQ:: Pie-Chart with frequency of Splicing-Types 

```{r splice_events_and_colours}

splice_events.mj <- c("alt3prime",
                      "alt5prime",
                      "alternate_first_exon",
                      "alternate_last_exon",
                      "alternative_intron",
                      "cassette",
                      "heatmap",
                      "junctions",
                      "multi_exon_spanning",
                      "mutually_exclusive",
                      "orphan_junction",
                      "p_alt3prime",
                      "p_alt5prime",
                      "p_alternate_first_exon",
                      "p_alternate_last_exon",
                      "tandem_cassette" )

##> create Colors-vector by : r-charts.com/colors/
my_events_colours <- c(
                  "alt3prime"              = "#FF1493",
                  "alt5prime"              = "#8B0A50",
                  "alternate_first_exon"   = "#7FFF00",
                  "alternate_last_exon"    = "#A2CD5A",
                  "alternative_intron"     = "#FFF0F5",
                  "cassette"               = "#E15759",
                  "heatmap"                = "#2F4F4F",
                  "junctions"              = "#1C86EE",
                  "multi_exon_spanning"    = "#F28E2B",
                  "mutually_exclusive"     = "#00EEEE",
                  "orphan_junction"        = "#9932CC",
                  "p_alt3prime"            = "#59A14F",
                  "p_alt5prime"            = "#FF3030",
                  "p_alternate_first_exon" = "#9BCD9B",
                  "p_alternate_last_exon"  = "#FFD700",
                  "tandem_cassette"        = "#EEE685" )

```

> for all samples, all splice-types of MAJIQ

```{r pie_chart_all_splice_types }

splice_types_sum.mj <- imap(majiq_mrg_tables,  
                                  \(sample_tab, sample_name) {
                                    df_se <- data.frame(t(table(factor(sample_tab$event_type))))
                                    colnames(df_se) <- c("Var1","Event_Type","Frequency")
                                    df_se <- df_se %>% select(-"Var1") 
                                    df_se
                                    ##! must always print the value to be exported outside imap!
                                  })
                                  
##> Sum over all samples : 
global_splice_sum.mj <- splice_types_sum.mj  %>% 
                      bind_rows(.id = "sample_name") %>% 
                      group_by(Event_Type) %>%  
                      summarise(Total_Number = sum(Frequency, na.rm = TRUE),.groups = "drop")
      
write_delim(global_splice_sum.mj,
        file="SpliceTypes_piecharts/Global_EventTypes_Sum_MAJIQ.txt",
        delim="\t",
        na="NA",
        col_names = TRUE)

event_types <- levels(as.factor(global_splice_sum.mj$Event_Type ))


global_splice_pct.mj <- global_splice_sum.mj  %>% 
                        mutate(
                          Percentage = Total_Number / sum(Total_Number) * 100,
                          label = paste0(round(Percentage, 1), "%","(",Total_Number ,")")
                          )

## Use a Palette to cover all events and be flexible for different number of slices in piechart
library(geomtextpath)

  
##> Visualise and label :
p <- ggplot(global_splice_pct.mj, aes(x = "", y = Total_Number, fill=Event_Type )) +
            geom_col(width = 1, stat="identity",colour = "white", linewidth = 0.4) +
            coord_polar(theta = "y", start=0) +
            scale_fill_manual(values = my_events_colours,
                          na.value = "grey70",
                          guide="none") +     #guide: for legend of "event_type"
        labs(
            title    = glue::glue(" MAJIQ-Event Types [All samples]"),
            subtitle = paste("Total events:", 
                      scales::comma(sum(global_splice_sum.mj$Total_Number))),
            fill     = "Event_Type") +
        theme(
              panel.background = element_blank(),
              plot.background = element_blank(),
              axis.text = element_blank(),
              axis.ticks = element_blank()) +
        # Percentage text 
        geom_text(aes(x = 1.2, label = paste0(round(Percentage, 1), "%")),
                  position = position_stack(vjust = 0.5),
                  size = 3.5,
                  colour = "#370335",
                  fontface = "bold") +
        # Labels text 
        geom_labelpath(data = global_splice_pct.mj,
                       position = position_stack(vjust = 0.5),
                        aes(x = 1.6, y = Total_Number, label = Event_Type),
                        size = 3.2,
                        colour = "#370335",
                        show.legend = FALSE); 

p

```


> Write a FUNCTION for Count + PieChart  -> MAJIQ & rMATS 

#### Regroup Splicing-Types to have only the basic splicing-events, comparable to rMATS-events 

-  Doc_MAJIQ : 
https://biociphers.bitbucket.io/majiq-docs/modulizer/event-types.html#putative-alternative-splice-sites-p-alt-5-3

 * "p_alt5prime"  | "alt5prime" | "alternate_first_exon" | "p_alternate_first_exon"  -> A5SS 
 
 * "p_alt3prime"  | "alt3prime"  | "alternate_last_exon" |"p_alternate_last_exon"    -> A3SS 
 
  \n(p_alt = putative alternative splice-site)
 
 * "alternative intron" = retained_intron      -> RI
 
 * "cassette" | "tandem_cassette"              -> SE
 
 * "multi_exon_spanning" , "heatmap"           -> "Others"
 
 * "junctions" | orphan_junction               -> Junctions
 
 * "mutually_exclusive"                        -> MXE
 
 
```{r function_calculate_splice_types_Sum}

count_splice_types4pie <- function(mrg_tab, tool) {
  
    ##> Create a list with frequency of each splice-event per sample : 
    ## imap: maps over each element of list and keeps names
    #' Input: list of data.frames from majiq/rmats results (merged splice_types in one table)
    #' Output : Data.frame with splice-event and frequency of each + Percentage of frequencies
    #' Output2: PieChart with Freq + Percentages 

    splice_types_sum <- imap(mrg_tab,  
                                  \(sample_tab, sample_name) {
                                    if ("event_type" %in% colnames(sample_tab)) {
                                      #change column to be as rmats:splice_event
                                      sample_tab <- dplyr::rename(sample_tab, "splice_event" ="event_type")
                                    }
                                    df_se <- data.frame(t(table(factor(sample_tab$splice_event))))
                                    colnames(df_se) <- c("Var1","Splice_Event","Frequency")
                                    df_se <- df_se %>% select(-"Var1") 
                                    df_se
                                    ##! must always print the value to be exported outside imap!
                                  })
                                  
    ##> Sum over all samples : 
    global_splice_sum <- splice_types_sum  %>% 
                          bind_rows(.id = "sample_name") %>% 
                          group_by(Splice_Event) %>%  
                          summarise(Total_Number = sum(Frequency, na.rm = TRUE),.groups = "drop")
          
    # write_delim(global_splice_sum.mj,
    #         file="SpliceTypes_piecharts/Global_SpliceEvents_Sum.txt",
    #         delim="\t",
    #         na="NA",
    #         col_names = TRUE)
    
    splice_events <- levels(as.factor(global_splice_sum$Splice_Event))
    
    ##> Regroup Splice-Types if they are more than known-rmats groups :
    if ( length(splice_events) > 7 ) { 
      
        global_splice_sum2 <- global_splice_sum %>% mutate(
                              Splice_Type = case_when(
                              Splice_Event %in% c("alt5prime", "alternate_first_exon",
                                                  "p_alt5prime", "p_alternate_first_exon") ~ "A5SS",
                              Splice_Event %in% c("alt3prime", "alternate_last_exon",
                                                  "p_alt3prime", "p_alternate_last_exon")  ~ "A3SS",
                              Splice_Event %in% c("cassette", "tandem_cassette")           ~ "SE",
                              Splice_Event %in% c("junctions", "orphan_junction")          ~ "Junctions",
                              Splice_Event %in% c("multi_exon_spanning", "heatmap")        ~ "Others",
                              Splice_Event == "mutually_exclusive"                         ~ "MXE",
                              Splice_Event == "alternative_intron"                         ~ "RI",
                              TRUE ~ NA   # last security-escape, all event not-listed
                            )) %>%
                            group_by(Splice_Type) %>%
                            summarise(Total_Number = sum(Total_Number), .groups = "drop")
        
         # change Splice_Group in factor for pie:
        global_splice_sum2 <- mutate(global_splice_sum2,
                                    Splice_Event = factor(Splice_Type))
        
        global_splice_sum <- global_splice_sum2 
        
     }
    
    global_splice_pct <- global_splice_sum  %>% 
                            mutate(
                              Percentage = Total_Number / sum(Total_Number) * 100,
                              label = paste0(round(Percentage, 1), "%","(",Total_Number ,")")
                              )

    ## Use a Palette to cover all events and be flexible for different number of slices in piechart
    library(paletteer) ## install.packages("paletteer")
    library(geomtextpath)
    # ! Attention to verify that values are not more than colours provided, here =18 !
    events_colours <- paletteer_d("ggsci::uniform_startrek") 

    ##> Visualize and label :
    p <- ggplot(global_splice_pct, aes(x = "", y = Total_Number, fill=Splice_Event )) +
                geom_col(width = 1, stat="identity",colour = "white", linewidth = 0.4) +
                coord_polar(theta = "y", start=0) +
                scale_fill_manual(values = events_colours,
                              na.value = "grey70",
                              guide="none") +     #guide: for legend of "event_type"
            labs(
                title    = glue::glue("{tool} Splicing Events [All samples]"),
                subtitle = paste("Total events:", 
                          scales::comma(sum(global_splice_sum$Total_Number))),
                fill     = "Splice_Event") +
            theme(
                  panel.background = element_blank(),
                  plot.background = element_blank(),
                  axis.text = element_blank(),
                  axis.ticks = element_blank()) +
            # Percentage text 
            geom_text(aes(x = 1.2, label = paste0(round(Percentage, 1), "%")),
                      position = position_stack(vjust = 0.5),
                      size = 3.5,
                      colour = "#370335",
                      fontface = "bold") +
            # Labels text 
            geom_labelpath(data = global_splice_pct,
                           position = position_stack(vjust = 0.5),
                            aes(x = 1.6, y = Total_Number, label = Splice_Event),
                            size = 3.2,
                            colour = "#370335",
                            show.legend = FALSE); 
    
    print(p)
    return(p)
  ## modif with "geom_labelpath" by : 
  ## https://stackoverflow.com/questions/74930665/r-fixing-labels-in-pie-chart

}

```

 Call Function for Count + Plot

```{r piechart_sumup_4majiq}

p.majiq <- count_splice_types4pie(majiq_mrg_tables,"MAJIQ")
p.majiq.filt <- count_splice_types4pie(majiq_mrg_tables.filt,"MAJIQ")

##> Save output in PNG:
dir.create("SpliceTypes_piecharts/",showWarnings = FALSE)
pdf("SpliceTypes_piecharts/Piechart_All_samples_all-splice-types_Majiq.pdf")
  print(p.majiq)
  print(p.majiq.filt)
dev.off()

    
```


```{r save_in_file,echo=T,eval=F}

###> Keep also in file :

write_delim(global_splice_pct.mj,
            file="SpliceTypes_piecharts/Global_SpliceGroups_Sum&Percentage.txt",
            delim="\t",
            na="NA",
            col_names = TRUE)

if (file.exists("SpliceTypes_piecharts/Global_SpliceGroups_Sum&Percentage.txt")){
  print("Saved in file: \'SpliceTypes_piecharts/Global_SpliceGroups_Sum&Percentage.txt\' ")
}

```


```{r piechart_No2_spliceGroups,echo=F,eval=F }

Group_events_colours <- c( "A5SS"      = "#FFC125",
                           "A3SS"      = "#FFEFDB",
                           "SE"        = "#FF7256",
                           "Junctions" = "#1C86EE",
                           "MXE"       = "#ADD8E6",
                           "Others"    = "#7CCD7C",
                           "NA"        = "grey70"  )


global_splice_sum2.mj <- global_splice_sum.mj %>% mutate(
                            Splice_Type = case_when(
                              Event_Type %in% c("alt5prime", "alternate_first_exon",
                                                "p_alt5prime", "p_alternate_first_exon") ~ "A5SS",
                              Event_Type %in% c("alt3prime", "alternate_last_exon",
                                                "p_alt3prime", "p_alternate_last_exon")  ~ "A3SS",
                              Event_Type %in% c("cassette", "tandem_cassette")           ~ "SE",
                              Event_Type %in% c("junctions", "orphan_junction")          ~ "Junctions",
                              Event_Type %in% c("multi_exon_spanning", "heatmap")        ~ "Others",
                              Event_Type == "mutually_exclusive"                         ~ "MXE",
                              Event_Type == "alternative_intron"                         ~ "RI",
                              TRUE ~ NA   # last security-escape, all event not-listed
                            )) %>%
                            group_by(Splice_Type) %>%
                            summarise(Total_Number = sum(Total_Number), .groups = "drop")
        
 # change Splice_Group in factor for pie:
global_splice_sum2.mj <- mutate(global_splice_sum2.mj,
                                Splice_Type = factor(Splice_Type))
        

global_splice_pct.mj <- global_splice_sum2.mj  %>% 
                            mutate(
                              Percentage = Total_Number / sum(Total_Number) * 100,
                              label = paste0(round(Percentage, 1), "%","(",Total_Number ,")")
                              )


##> Visualise and label :
p2 <- ggplot(global_splice_pct.mj, aes(x = "", y = Total_Number, fill=Splice_Type )) +
            geom_col(width = 1, stat="identity", colour = "white", linewidth = 0.4) +
            coord_polar(theta = "y", start=0) +
            scale_fill_manual(values = Group_events_colours,
                          na.value = "grey70",
                          guide="none") +     #guide: for legend of "event_type"
            labs(
                  title    = "MAJIQ - Splicing Events [All samples]",
                  subtitle = paste("Total events:",
                            scales::comma(sum(global_splice_sum2.mj$Total_Number))),
                  fill  = "Splice_Type") +
            theme(
                panel.background = element_blank(),
                plot.background = element_blank(),
                axis.text = element_blank(),
                axis.ticks = element_blank()) +
          # Percentage text
          geom_text(aes(x = 1.2, label = paste0(round(Percentage, 1), "%")),
                    position = position_stack(vjust = 0.5),
                    size = 4,
                    colour = "#370335",
                    fontface = "bold") +
          # Labels text
          geom_labelpath(data = global_splice_pct.mj,
                        position = position_stack(vjust = 0.5),
                        aes(x = 1.6, y = Total_Number, label = Splice_Type),
                        size = 3,
                        colour = "#370335",
                        show.legend = FALSE);

print(p2)

##> NOTE : 
##' x=1 of geom_labelpath()| geom_text() locates the label more or less close to center.
##' x < 1 : closer to center, x >1 : further from center to periphery


png("SpliceTypes_piecharts/Piechart_All_samples_SpliceGroups_withPercentage_Majiq.png")
  print(p2)
dev.off()

# if (file.exists("SpliceTypes_piecharts/Piechart_All_samples_SpliceGroups_withPercentage_Majiq.png")){
#   print("Pie Chart saved in png:\n \'SpliceTypes_piecharts/Piechart_All_samples_SpliceGroups_withPercentage_Majiq.png\'")
# }

```

#### rMATS - PieChart of Splicing Events on final_merged_tables 

```{r rmats_pieChart_on_splice_events}

## par "rmats_dir/merged_tables" -> rmats_mrg_tables

# splice_types_sum.rmats <- imap(rmats_mrg_tables,  
#                               \(rm_tab, sample_name) {
#                                 rm_events <- data.frame(t(table(factor(rm_tab$splice_event))));
#                                 colnames(rm_events) <- c("Var1","Splice_Event","Frequency")
#                                 rm_events ##! must always print the value to be exported outside imap!
#                               })
#                               
# ##> Sum over all samples : 
# global_splice_sum.rmats <- splice_types_sum.rmats  %>% 
#                       bind_rows(.id = "sample_name") %>% 
#                       group_by(Splice_Event) %>%  
#                       summarise(Total_Number = sum(Frequency, na.rm = TRUE),.groups = "drop")
#                       
# ##> with Percentage :
# 
# global_splice_pct.rmats <- global_splice_sum.rmats  %>% 
#                       mutate(
#                         percentage = Total_Number / sum(Total_Number) * 100,
#                         label = paste0(round(percentage, 1), "%","(",Total_Number ,")")
#                       )

# library(geomtextpath)
# 
# p.rmats <- ggplot(global_splice_pct.rmats, 
#                   aes(x = "", y = percentage, fill=Splice_Event )) +
#                   geom_col(width = 1, stat="identity", colour = "white", linewidth = 0.4) +
#                   coord_polar(theta = "y", start=0) +
#                   scale_fill_manual(values = events_colours,
#                                     na.value = "grey70",
#                                     guide="none") +     #guide: for legend of "event_type"
#                   labs(
#                       title    = "rMATS-Splicing Events [All samples]",
#                       subtitle = glue::glue("Total events: {sum(global_splice_pct.rmats$Total_Number)}"),
#                       fill     = "Splice_Event") +
#                   theme(
#                         panel.background = element_blank(),
#                         plot.background = element_blank(),
#                         axis.text = element_blank(),
#                         axis.ticks = element_blank()) +
#                   # Percentage text INSIDE the slices
#                   geom_text(aes(x = 0.95, label = paste0(round(percentage, 1), "%")),
#                             position = position_stack(vjust = 0.5),
#                             size = 4,
#                             colour = "#030303",
#                             fontface = "bold") +
#                   geom_labelpath(data = global_splice_pct.rmats,
#                                  position = position_stack(vjust = 0.5),
#                                   aes(x = 1.6, y = percentage, label = Splice_Event),
#                                   size = 3,
#                                   show.legend = FALSE); 
# 
# p.rmats

```


```{r call_function_pieChart_rmats}

p.rmats <- count_splice_types4pie(rmats_mrg_tables,"rMATS")
p.rmats.filt <- count_splice_types4pie(rmats_mrg_tables.filt,"rMATS")


dir.create("SpliceTypes_piecharts/",showWarnings = FALSE)
pdf("SpliceTypes_piecharts/Piechart_All_samples_all-splice-types_rMATS.pdf",onefile = T)
  print(p.rmats)
  print(p.rmats.filt)
dev.off()


```

