---
title: "Pathways_of_spliced_genes"
author: "Maria Kondili"
date: "2026-06-09"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libs}

library(tidyverse)
library(clusterProfiler)
library(fgsea)

```

#### Rename column of DPSI,necessary for gsea-function

```{r rename_dpsi_column_for_gsea}

majiq_mrg_tables.renam <- map(majiq_mrg_tables, 
                              ~(.x <- rename(.x, 
                                             "dPSI" = "grp2-grp1_median_dpsi",
                                             "geneSymbol" = "gene_name")
                            ))

```

```{r function: run_gsea_analysis.R }

#input_data <- majiq_mrg_tables$C002I3K$grp2-grp1_median_dpsi
# majiq_mrg_tables$C002I3K %>% mutate(`grp2-grp1_median_dpsi` = "dpsi")


.run_gsea_analysis <- function( input_data,sampleID, myFDR, myNES){

  # test : input_data <- majiq_mrg_tables.filt.renam$C002I3M
  gsea_genes <- input_data$dPSI # instead of FoldChange
  names(gsea_genes) <- input_data$geneSymbol # changed as rMATS column
  #gsea_genes <- na.omit(gsea_genes)
  #--> genes with FC="NA".IF in not any NA,will return original input
  
  dups <- duplicated(names(gsea_genes))
  ## if there are duplicates (dups ==TRUE ):
  if ( any(dups) ) {
    gsea_genes <- gsea_genes[-which(dups)]
  }
  gsea_genes <- sort(gsea_genes, decreasing=T)
  
  #! sorting removes also NAs in values, but not in names:
  #> remove any genes with name=NA 
  if ( any(is.na(names(gsea_genes))) ) {
    
      gsea_genes <- gsea_genes[- which(is.na(names(gsea_genes)))]
  }
  
  ###
  ### FGSEA
  ###
  
  pathw_dir="~/MYO_WORKSPACE/myScriptsBank/MSigDB_Hallmark_Pathways/2023/"
  
  set.seed(42)
  
  #----2023-----
  #gmt_listfiles <- list.files(pathw_dir)
  c2 <- gmtPathways(paste0(pathw_dir, "c2.all.v2023.2.Hs.symbols.gmt"))
  c5 <- gmtPathways(paste0(pathw_dir, "c5.all.v2023.2.Hs.symbols.gmt"))
  hm <- gmtPathways(paste0(pathw_dir, "h.all.v2023.2.Hs.symbols.gmt"))
  ##ms <- gmtPathways(paste0(pathw_dir, "msigdb.v2023.2.Hs.symbols.gmt")) -> duplicate terms
  
  
  gseaRes.c2 <- fgsea(c2, gsea_genes, minSize=15, maxSize=1000)
  gseaRes.c5 <- fgsea(c5, gsea_genes, minSize=15, maxSize=1000)
  gseaRes.hm <- fgsea(hm, gsea_genes, minSize=15, maxSize=1000)
  
  ## Filter with myFDR :
  gseaRes <- dplyr::bind_rows( gseaRes.c2, gseaRes.c5,gseaRes.hm ) %>%
               dplyr::filter(padj < myFDR ) %>% arrange(desc(NES))
  
  ## Reduce Redundant
  # pathway2name <-  gseaRes[order(padj), setNames(pathway, pathway)]
  # reduced_gsea <- eliminatePathways(universe=names(gsea_genes),
  #                                 pathways=list(c2,c5,hm),
  #                                 pathway2name=pathway2name,
  #                                 isNonRandomPval=fgseaIsNonRandomPval(list(c2,c5,hm), gsea_genes, nperm=5000))
  #
  
  ## Filter with myNES:
  gseaRes2show <- subset(gseaRes, abs(NES)> myNES )
  
  ## Concatenate "LEADING_EDGE" Genes in one cell
  gseaRes$leadingEdge <- unlist(map(gseaRes$leadingEdge , ~paste(.x ,collapse=",")))
  
  g <- ggplot(gseaRes2show, aes( reorder(pathway, NES),NES)) +
              geom_col(aes(fill = padj)) +
              coord_flip() +
              labs(x="Pathway", 
                   y="Normalized Enrichment Score",
                   title= glue::glue("{sampleID},All signif Pathways (|NES|>{myNES})")
                   ) +
              theme(axis.text = element_text(size = 5)) +
              theme(axis.title = element_text(size = 11))
  print(g)
  
  return(gseaRes)
}

  
```


```{r test_func_1sample } 

gseaRes.C002I3M <- .run_gsea_analysis(majiq_mrg_tables.renam$C002I3M,
                                      sampleID="MajiQ::C002I3M",
                                      myFDR=0.05,
                                      myNES=0 )


gseaRes.C002I3Z <- .run_gsea_analysis(majiq_mrg_tables.renam$C002I3Z,
                                      sampleID="MajiQ::C002I3Z",
                                      myFDR=0.05,
                                      myNES=0 )

```

#### MAJIQ spliced genes - GSEA 

```{r run_func_majiq_gsea}

sampleIDs <- names(majiq_mrg_tables.renam)
gsea_pathw_per_sample.mj    <- map2(.x = majiq_mrg_tables.renam,
                                    .y = sampleIDs,
                                    ~.run_gsea_analysis(.x,
                                                        sampleID=.y,
                                                        myFDR=0.05,
                                                        myNES=0))


null <- map(gsea_pathw_per_sample.mj, ~print(nrow(.x)))

```

#### rMATS spliced genes - GSEA

```{r run_func_rmats_gsea}

sample_names <- names(rmats_mrg_tables)
gsea_pathw_per_sample.rmats <- map2(.x = rmats_mrg_tables,
                                    .y = sample_names,
                                   ~.run_gsea_analysis(.x,
                                                       sampleID=.y,
                                                       myFDR=0.05,
                                                       myNES=0))

null <- map(gsea_pathw_per_sample.rmats, ~print(nrow(.x)))

## test 1sample: 

.run_gsea_analysis(rmats_mrg_tables$C002I3Z,sampleID="C002I3Z", myFDR=0.05,myNES=0)

```

!!! 
Message d'avis :
Dans preparePathwaysAndStats(pathways, stats, minSize, maxSize, gseaParam,  :
  There are ties in the preranked stats (87.84% of the list).
The order of those tied genes will be arbitrary, which may produce unexpected results.


####  ORA - analysis for same genes 

clusterProfiler::enricher(
   gene,
   pvalueCutoff = 0.05,
   pAdjustMethod = "BH",
   universe = NULL,
   minGSSize = 10,
   maxGSSize = 500,
   qvalueCutoff = 0.2,
   gson = NULL,
   TERM2GENE, # --> two columns,
   TERM2NAME = NA
 )



-------r- BIOAPI (https://cran.r-project.org/web/packages/rbioapi/vignettes/rbioapi_do_enrich.html)----


```{r}
#install.packages("rbioapi")
library(rbioapi)

# reactome <- rba_reactome_analysis(input = covid_critical)

```

> MAJIQ Genes

```{r}
#geneids_K <- majiq_mrg_tables.renam$C002I3K$gene_id
enrichr_libs <- rba_enrichr_libs(organism = "human")

enrichr_libs #--> choose which to input @"gene_set_library"
#..
#..
#..

```


```{r enrichr_for_majiq_genes }

enrichr_GO_allsamples.mj <- map(majiq_mrg_tables.renam, 
                             ~rba_enrichr(gene_list =.x$geneSymbol,
                                          organism = "human",
                                          gene_set_library = c("GO_Biological_Process_2023",
                                                               "GO_Molecular_Function_2023",
                                                               "GO_Cellular_Component_2023"),
                                          regex_library_name = FALSE,
                                          background_genes = NULL,
                                          progress_bar = FALSE)
)
                            


```

> rMATS

```{r enrichr_for_rmats_genes}

## show how many genes for each sample: 
map(rmats_mrg_tables, ~print(length(unique(.x$geneSymbol))))
                                          

enrichr_GO_allsamples.rm <- map(rmats_mrg_tables, 
                             ~rba_enrichr(gene_list =.x$geneSymbol,
                                          organism = "human",
                                          gene_set_library = c("GO_Biological_Process_2023",
                                                               "GO_Molecular_Function_2023",
                                                               "GO_Cellular_Component_2023"),
                                          regex_library_name = FALSE,
                                          background_genes = NULL,
                                          progress_bar = FALSE)
)

```

### Compare :

> 1. FILTER

```{r filter_GOterms_results}
## Function:
filter_enrichr_res <- function(sample_enrichr){
  
  BP.filt <- sample_enrichr$GO_Biological_Process_2023 %>% dplyr::filter(Adjusted.P.value < 0.05)
  MF.filt <- sample_enrichr$GO_Molecular_Function_2023 %>% dplyr::filter(Adjusted.P.value < 0.05)
  CC.filt <- sample_enrichr$GO_Cellular_Component_2023 %>% dplyr::filter(Adjusted.P.value < 0.05)
  
  all_GO_enrichr <- bind_rows(BP.filt,MF.filt,CC.filt)
  
  return(all_GO_enrichr)
  
}

## Call Function :
all_GO_filt.mj <- map(enrichr_GO_allsamples.mj, ~filter_enrichr_res(.x))
all_GO_filt.rm <- map(enrichr_GO_allsamples.rm, ~filter_enrichr_res(.x))


```
> 1.b Show Number of GO-terms per sample :

```{r nb_of_terms_found}


nb_terms.mj <- map(all_GO_filt.mj, ~length(.x$Term))
nb_terms.rm <- map(all_GO_filt.rm, ~length(.x$Term))

##!!! GO-terms from MAJIQ are all Zero except "M"-sample
##!!! GO-terms from rMATS-results are > 1000 for all samples !
```

> 2. find common GO-terms

```{r func_find_common_go_terms}

## Function: 
find_common_GOterms <- function(GO_tab_filt.mj , GO_tab_filt.rm,sample_name ){
  
  ##> TABLE_1
    majiq_terms <- GO_tab_filt.mj[[sample_name]] %>%
                    pull(Term) %>%
                    unique() %>%
                    na.omit()
    ## TABLE_2
    rmats_terms <- GO_tab_filt.rm[[sample_name]] %>%
                    pull(Term) %>%
                    unique() %>%
                    na.omit()
    
   
    common_terms <- intersect(majiq_terms , rmats_terms )
    return(common_terms)
  
}

## Call Function :
common_GOterms <- map(names(all_GO_filt.rm), 
                      ~find_common_GOterms(all_GO_filt.mj,
                                           all_GO_filt.rm,
                                            .x))
  
 
```

--> common_GOterms 

[[1]]
character(0)

[[2]]
character(0)

[[3]]
[1] "Striated Muscle Contraction (GO:0006941)"

[[4]]
character(0)

[[5]]
character(0)

[[6]]
character(0)

[[7]]
character(0)

[[8]]
 [1] "Organelle Organization (GO:0006996)"                                        
 [2] "Endosomal Transport (GO:0016197)"                                           
 [3] "Vesicle-Mediated Transport (GO:0016192)"                                    
 [4] "Regulation Of Focal Adhesion Assembly (GO:0051893)"                         
 [5] "Membrane Organization (GO:0061024)"                                         
 [6] "Protein Localization (GO:0008104)"                                          
 [7] "post-Golgi Vesicle-Mediated Transport (GO:0006892)"                         
 [8] "Cell-Substrate Junction Assembly (GO:0007044)"                              
 [9] "Regulation Of Cell-Substrate Junction Assembly (GO:0090109)"                
[10] "Cellular Response To Insulin Stimulus (GO:0032869)"                         
[11] "Cytosolic Transport (GO:0016482)"                                           
[12] "Positive Regulation Of Apoptotic Process (GO:0043065)"                      
[13] "Glycerophospholipid Biosynthetic Process (GO:0046474)"                      
[14] "Intracellular Protein Transport (GO:0006886)"                               
[15] "Regulation Of Anatomical Structure Morphogenesis (GO:0022603)"              
[16] "Regulation Of Stress Fiber Assembly (GO:0051492)"                           
[17] "Focal Adhesion Assembly (GO:0048041)"                                       
[18] "Golgi Vesicle Transport (GO:0048193)"                                       
[19] "Actomyosin Structure Organization (GO:0031032)"                             
[20] "Positive Regulation Of Cell Migration (GO:0030335)"                         
[21] "Retrograde Transport, Endosome To Golgi (GO:0042147)"                       
[22] "Cytoskeleton Organization (GO:0007010)"                                     
[23] "Endomembrane System Organization (GO:0010256)"                              
[24] "Protein Modification Process (GO:0036211)"                                  
[25] "Protein Transport (GO:0015031)"                                             
[26] "Regulation Of Cell Migration (GO:0030334)"                                  
[27] "Endocytosis (GO:0006897)"                                                   
[28] "Response To Insulin (GO:0032868)"                                           
[29] "Peptidyl-Threonine Modification (GO:0018210)"                               
[30] "Peptidyl-Serine Phosphorylation (GO:0018105)"                               
[31] "Negative Regulation Of Protein Localization To Cell Periphery (GO:1904376)" 
[32] "Negative Regulation Of Protein Localization To Plasma Membrane (GO:1903077)"
[33] "Lamellipodium Organization (GO:0097581)"                                    
[34] "Endoplasmic Reticulum Organization (GO:0007029)"                            
[35] "Vesicle-Mediated Transport To The Plasma Membrane (GO:0098876)"             
[36] "Positive Regulation Of Stress Fiber Assembly (GO:0051496)"                  
[37] "Plasma Membrane Bounded Cell Projection Assembly (GO:0120031)"              
[38] "Peptidyl-Serine Modification (GO:0018209)"                                  
[39] "Peptidyl-Threonine Phosphorylation (GO:0018107)"                            
[40] "Golgi To Plasma Membrane Transport (GO:0006893)"                            
[41] "Insulin Receptor Signaling Pathway (GO:0008286)"                            
[42] "Receptor-Mediated Endocytosis (GO:0006898)"                                 
[43] "Phosphorylation (GO:0016310)"                                               
[44] "Golgi To Endosome Transport (GO:0006895)"                                   
[45] "Negative Regulation Of Cell-Substrate Junction Organization (GO:0150118)"   
[46] "Negative Regulation Of Focal Adhesion Assembly (GO:0051895)"                
[47] "Negative Regulation Of Protein Localization To Membrane (GO:1905476)"       

[[9]]
character(0)

[[10]]
character(0)

[[11]]
character(0)

[[12]]
character(0)

[[13]]
character(0)

[[14]]
character(0)

[[15]]
character(0)

[[16]]
character(0)

[[17]]
[1] "Positive Regulation Of Cell-Substrate Junction Organization (GO:0150117)"
