---
title: "Filter signif ASE with Muscle-Disease Genes"
author: "Maria Kondili"
date: "2025-10-15"
output: 
 html_document: default
editor_options:
  chunk_output_type: console
---

Objective: Detect the A.S.E. with highest dPSI and lowest FDR, from the merged_ASE_FDR_sorted tables and also that fall over a muscle-disease gene (from published "gene-tables", Bonne & Rivier )

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(glue))

library(readxl)
library(dplyr)
library(stringr)
library(purrr)
library(tibble)
```


```{r}

#work_dir="/shared/projects/splicing_events_detection_rare_diseases/EPIS_SAV/rMATS/"

work_dir="~/MYO_WORKSPACE/PROJECTS/EPIS_SAV-JohnRendu/rMATS/Downstream_Analysis/"
setwd(work_dir)
load(".RData")

#downstream_outdir = paste0(work_dir,"Outputs/one_group_stat_off/Downstream_Outputs/")
downstream_outdir = paste0(work_dir,"../Outputs/one_group_stat_off/Downstream_Outputs/")

dir.create(downstream_outdir, showWarnings = F)

#mrg_filt_outdir <- paste0(downstream_outdir,"Merged_Filtered_ASEs/") # created inside sample "final"
#dir.create(mrg_filt_outdir ,showWarnings = FALSE)

```


```{r samples }

samples <- c("4F","4G","44","45","47","K","L","M","N","P","Q","R","S","U","W","Y","Z")

```

### Filter splicing events by dPSI & FDR

```{r dir_filtered}

## create new folder to save filtered-rmats-results:
filtres_dir <- paste0(downstream_outdir, "filt_by_dPSI_FDR_GeneTables/" ) 

dir.create(filtres_dir, showWarnings = FALSE)

```

```{r filter_signif_mrg_ase,echo=F,eval=F}

for (s in samples) {
  
  # Create output folder to save merged table
  final_outdir  <- paste0(downstream_outdir,"final_",s,"_vs_all/")
  
  mrg_ase_file <- paste0(final_outdir,"merged_modif_ASE_sample",s,"_FDRsorted.tsv")
  
  mrg_ase <- read_delim(mrg_ase_file,delim = "\t", col_names = TRUE, show_col_types = FALSE)
  
  mrg_ase.filt <- mrg_ase %>% dplyr::filter(FDR <=0.05 & dPSI >= 0.5)
  
  print(glue::glue("~~> {nrow(mrg_ase.filt)} Significant events found !"))

  head(mrg_ase.filt)
  
  write_delim(mrg_ase.filt,
              paste0(filtres_dir, "merged_filt_modif_ASE_sample",s,".txt"),
              delim = "\t",col_names = TRUE)

}

```

#### Read MUSCLE-GENE-TABLES --> create a merged table for all diseases

```{r read_geneTables}

##1/ Read Tables and  add disease in a column --> merge tables to one

gene_tables_dir <- paste0(work_dir,"../GeneTables_Bonne-Rivier/")

mu_genetables <- list.files(gene_tables_dir, full.names = T)

# names(mu_genetables) <- basename(mu_genetables) 
#--> to be used if map_(.id="SourceFile") for detecting where each line of gene comes from
## not needed since I added column of disease 

diseases <- purrr::map_chr( mu_genetables, \(x) sub(".xlsx","", basename(x)) )


read_gene_table <- function(genetable) {
  
  ## Input: filename of excel gene-table
  ## Output: a tibble with modif.column names and disease_name new column
  
  genetab <- readxl::read_xlsx(genetable, skip = 1, col_names = TRUE)
  
  gene_col <- names(genetab)[str_detect(names(genetab), "Gene symbol")] 
  
  disease <- sub(".xlsx","", basename(genetable))
  
  genetab_modif <- genetab %>% 
                    rename("GeneSymbol" = all_of(gene_col),
                           "Item_in_genetable"=`Item in the gene table`) %>%
                    mutate(Disease_name = disease)
                           
  return(genetab_modif)
  
}

```

#### Merge gene-tables 

```{r merge_genetables }

##from chatGPT: 
mrg_genetables <- purrr::map_dfr(mu_genetables, read_gene_table, .id = "SourceFile")

### Concatenate Diseases for each gene, to have unique Genes : 
mrg_genetables_collapsed <- mrg_genetables %>%
                              group_by(GeneSymbol) %>%
                              summarise(Disease = paste(unique(Disease_name), collapse = "; ")) %>%
                              ungroup()

##write the merged-table in a file :
readr::write_delim(mrg_genetables_collapsed,
                   file = paste0(gene_tables_dir,"Merged_GeneTables_collapsed_diseases.txt"),
                   delim = "\t",
                   col_names = TRUE,
                   na = "NA")
                  

```

### Filter dy dPSI & FDR after chosing the disease-genes 

Better calibration of output-genes when each filter is applied one by one and disease-genes at the end.

```{r}

filter_ase_by_gene_and_stats <- function(rmats_dir, genetable, my_fdr=0.05, my_dpsi= 0.5){
  
      s <-  sub("_vs_all","",sub("final_", "",basename(rmats_dir)) )
      print(glue::glue("\n\n~~>Sample treated: {s}"))
      
      print(glue::glue("!Cut-offs chosen:FDR <= {my_fdr}, |dPSI| >= {my_dpsi}"))
  
      ## read the merged-rmats-output:
      rmats_ase <- paste0(rmats_dir,"/merged_modif_ASE_sample",s,"_FDRsorted.tsv")
      
      ase_tab <- read_delim(rmats_ase, delim = "\t", col_names = TRUE, show_col_types = F )
      
    
      ## Filter by stat:
      ase_tab.filt <-  ase_tab %>% dplyr::filter(FDR <= my_fdr )
      
      print(glue::glue("#{length(ase_tab.filt$geneSymbol)} Significant events found for these genes !"))
  
     
      ## Filter by dPSI :
      ase_tab.filt2 <-  ase_tab.filt %>% dplyr::filter( abs(dPSI) >= my_dpsi)
      
      print(glue::glue("#{length(ase_tab.filt2$geneSymbol)} Highly spliced & signif. events found for these genes !"))

      
      ## Find Muscle-disease Genes (unique after collapsing) and keep Disease_name too :
      
      ase_tab.filt3 <- ase_tab.filt2 %>% filter(geneSymbol %in% mrg_genetables$GeneSymbol )
      
      # ase_tab.filt <- ase_tab %>% inner_join(genetable %>% 
      #                                         dplyr::select(GeneSymbol,Disease),
      #                                                by=c("geneSymbol" = "GeneSymbol") )
      # 
      ##NOTE 1: inner_join() keeps unique from "ase_tab" and "genetable"
      ##         left_join() keeps all genes/splicing from ase_tab
      ##NOTE 2 : adding "Disease" didn't help in well filtering the ase-tables, so only simple filtering of genes
      
      print(glue::glue("#{length(ase_tab.filt3$geneSymbol)} genes found associated to a disease,among the signif/ly spliced!\n"))
      
      
      return(ase_tab.filt3)
}

```

#### CALL function:

```{r call_function}

## create new folder to save filtered-rmats-results:

filtres_dir <- paste0(downstream_outdir, "filt_by_dPSI_FDR_GeneTables/" ) ## and Disease-Gene! 

final_outdirs <- list.files(paste0(downstream_outdir,"final_rmats_merged/"), pattern="final_*", full.names = TRUE)


for (rmats_dir in final_outdirs){
  
    s <-  sub("_vs_all","",sub("final_", "",basename(rmats_dir)) )
    ase_final.filt <- filter_ase_by_gene_and_stats(rmats_dir, mrg_genetables_collapsed,
                                                   my_fdr = 0.05, my_dpsi = 0.2)
    
    ##> Ignore the Disease,bcs genes resting, are 0 or very few with inner/left_join
    ## ase_final.filt  <- ase_final.filt %>% dplyr::relocate(c("Disease"), .before=c("chr"))
    
    
    ase_final.filt  <- ase_final.filt %>% dplyr::relocate(c("alternative_exon_coordinates","PValue", "FDR","dPSI"),
                                                          .before=c("IncFormLen"))

    ## remove non-necessary columns:
    ase_final.filt <- ase_final.filt %>% dplyr::select(- c(matches("IncLevel*"),"RowID"))
    
    print(ase_final.filt[1:4,1:6])
    
    readr::write_delim(ase_final.filt,
                      file=paste0(filtres_dir, "merged_filt_modif_ASE_sample",s,".txt"),
                      delim = "\t",col_names=TRUE,
                      append=FALSE)
     
}


```

Save all txt in Excel 

```{r save_in_excel}

library(openxlsx)

final_out_list <- list.files(filtres_dir, pattern="*.txt", full.names = TRUE)

df_list <- map(final_out_list, ~read_delim(.x))

names(df_list) <- samples
resu
write.xlsx(df_list, file=paste0(downstream_outdir,"KONTILI_InstitutMyologie_rMATS_final_lts_17patients.xlsx"))


```

Count number of Splicing Events on final results --> Pie chart

```{r count_events_types }

#read.xlsx(paste0(downstream_outdir,"KONTILI_InstitutMyologie_rMATS_final_results_17patients.xlsx"),sheet = names(df_list)??)

splice_events <- c("A3SS", "A5SS", "RI", "SE", "MXE")

##> Create a list with frequency of each splice-event per sample : 

splice_types_sum <- imap(df_list,  # maps over each element and keeps name 
                    \(li, sample_name) {
                      df <- data.frame(t(table(factor(li$splice_event, levels = splice_events))))
                      colnames(df) <- c("Var1","Splice_Event","Frequency")
                      df ##! must always print the value to be exported outside imap!
                     }
                    )

##> Sum over all samples : 

library(dplyr)
library(purrr)

global_splice_sum <- splice_types_sum |>
                      bind_rows(.id = "sample_name") |>
                      group_by(Splice_Event) |> 
                      summarise(Total_Number = sum(Frequency, na.rm = TRUE),.groups = "drop")
                      
                      


# global_splice_sum
# # A tibble: 5 × 2
#   Splice_Event Total_Number
#   <fct>               <int>
# 1 A3SS                   22
# 2 A5SS                    4
# 3 RI                     27
# 4 SE                     46
# 5 MXE                    11


##> Pie Chart : 
#---> old 
# pie(global_splice_sum$Total_Number,
#     labels=global_splice_sum$Splice_Event,
#     main="Distribution of Splicing Events")

##---> new 
library(ggplot2)
library(dplyr)

ggplot(global_splice_sum, aes(x = "", y = Total_Number, fill = Splice_Event)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  geom_text(aes(label=Total_Number), 
            position = position_stack(vjust = 0.5),
            size = 4)+
  labs(
    title = "Distribution of Splice Events over all patients",
    fill = "Splice Event") +
  theme_void()

```

Show Pie Chart with Percentages 

```{r add_percentage_in_sums}

#> add column to final tibble: 
global_splice_pct <- global_splice_sum |>
  mutate(
    percentage = Total_Number / sum(Total_Number) * 100,
    label = paste0(round(percentage, 1), "%")
  )

##> plot pie-chart with percentage 
ggplot(global_splice_pct, aes(x = "", y = percentage, fill = Splice_Event)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5),
            size = 4) +
  theme_void() +
  labs(title = "Distribution of Splice Events over all patients")
```

