---
title: From rMATS output to heatmap of diff splicing events on EPIS-SAV samples
author: "Maria Kondili"
date: "12/09/2025"
output:
  html_document: default
editor_options:
  chunk_output_type: console
---

Objective: Use rMATS output & Merge all events-types & Create heatmap of diff splicing events

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(glue))

```

```{r input_and_output_dir}

#work_dir="/shared/projects/shmbnl_rnaseq/4W_et_10W/rMATS_Splicing/"
work_dir="/shared/projects/splicing_events_detection_rare_diseases/EPIS_SAV/rMATS/"
load(".RData")

stat_outdir <- paste0(work_dir,"Outputs/one_group_stat_off/output_all_statoff/Stats_out/")
downstream_outdir <- paste0(work_dir,"Outputs/one_group_stat_off/","Downstream_Outputs/")

dir.create(downstream_outdir, showWarnings = F)

samples <- c( "44", "45","47","4F","4G","K","L","M","N","P","Q","R","S","U","W","Y","Z" ) 

## Create output folder to save merged table
for (s in samples) {
  final_outdir <- paste0(downstream_outdir,"final_",s,"_vs_all/")
  dir.create(final_outdir, showWarnings = F)

}


```

Modify rMATS tables with custom exon-coordinates (upstream/spliced/downstream positions)

```{r add_exon_info, message=FALSE, warning=FALSE}

functions_dir <- paste0(work_dir,"Downstream_analysis/functions/")

source(paste0(functions_dir,"add_exonLength_and_Coord.R"))

splice_events <- c("A3SS", "A5SS","MXE", "RI", "SE")
## output in same folder but with suffix "_exonCoord"


##> Read Output of STAT-step and add exon-length and coordinates 

for (s in samples) {
  cat("\n~~>Sample ", s,"is analysed now...\n")
  resdir_s <- paste0(stat_outdir,s,"_vs_all/") 
  final_outdir_s <- paste0(downstream_outdir,"final_",s,"_vs_all/")
  
  for (se in splice_events) {
      cat("~~> Launching event :",se,"\n")

      add_exonLen_and_coord(splice_event=se,
                            tab_suffix=".MATS.JCEC.txt",
                            in_dir  = resdir_s,
                            out_dir = final_outdir_s)
  }

  cat("New tables can be found here:\n")
  print(list.files(final_outdir_s ))
  
}

```

Merge Events in one table, with a given order of columns

```{r merge_events }

source(paste0(functions_dir,"/merge_spliceTypes_and_save.R"))

new_suffix = ".JCEC_PerRepl_exonCoord.txt"

inc_level_all_others <- paste("IncLevel2_r", seq(1,16), sep="")
  
colsOI <- c("GeneID",	"geneSymbol","chr","strand",
            "IncFormLen", "SkipFormLen","ExonLength",
            "IncLevel1_r1", inc_level_all_others,	
            "splice_event", "IncLevelDifference",
            "alternative_exon_coordinates","PValue","FDR" )

```

MERGE

```{r merge_output_events_per_sample }

for (s in samples) {
  #s="4F", "4G" #-> empty !!
  cat("\n~~> Outputs of Sample ",s,"will be merged now...\n" )
  
  final_outdir_s <- paste0(downstream_outdir,"final_",s,"_vs_all/")
  
  ase_mrg <- merge_spliceTypes_and_save(splice_events, 
                                        new_suffix, colsOI, 
                                        indir=final_outdir_s) 
                                      
  ase_mrg <- rename(ase_mrg,"dPSI"="IncLevelDifference")
  colnames(ase_mrg) <- gsub("IncLevel1_*", paste0("IncLevel_",s,"_"),colnames(ase_mrg))
  
  ase_mrg$RowID <- rownames(ase_mrg)
  rownames(ase_mrg) <- NULL

  ase_mrg %>%  glimpse

  mrg_filename=paste0("merged_modif_ASE_sample",s,"_FDRsorted.tsv")
  # write.table( ase_mrg[order(ase_mrg$FDR,decreasing = F),],
  #            paste0(final_outdir_s, mrg_filename),
  #            sep="\t", col.names = T,row.names = F,quote=F )
  
  readr::write_delim(x = ase_mrg[order(ase_mrg$FDR,decreasing = F),],
                    file = paste0(final_outdir_s, mrg_filename),
                     delim= "\t",col_names = TRUE,quote = NULL)


}

```

PLOT Heatmap with significantly spliced events 

```{r plot_heatmap_function,eval=F}

plot_heatmap_signif_splice <- function(ase, fdr.co, psi.co, pair, inc_level_cols= c(8:13)) {
  
    ase.filt <- subset(ase, (FDR < fdr.co &  abs(dPSI) > psi.co) )
    
    
    ###  z.gene =  (E(g) - Mean(g.replicates) ) / M.A.D (g.replicates), mad=median.absolute.deviation
    x <- ase.filt[,inc_level_cols]
    s1 = strsplit(pair, "_vs_")[[1]][1]
    s2 = strsplit(pair, "_vs_")[[1]][2]
    colnames(x) <- c( paste(s1,"_", c("r1","r2","r3"),sep=""),
                      paste(s2,"_",c("r1","r2","r3"),sep=""))
    
    
    cal_z_score <- function(x){
      (x - mean(x)) / sd(x)
    }
    x_norm <- t(apply(x, 1, cal_z_score))
    # x_norm <- x_norm[-which(apply(x_norm, 2, is.na)) ,]
    
    #Rowv=as.dendrogram(hcl_row),tree=ct))
    library(gplots)
    library(RColorBrewer)
    
   
    heatmap.2(x_norm, 
              Rowv=TRUE,
              Colv=F,
              dendrogram ='row',
              scale="row", na.rm=TRUE,
              trace='none',
              labCol=colnames(x),
              col=rev(colorRampPalette(brewer.pal(10, "RdBu"))(256)),
              margins=c(9,2),
              cexCol=1.3,
              srtCol=45,
              labRow="",
              offsetCol=0,
              key=TRUE,
              keysize = 1.5,
              density.info="density",
              key.title="Color range",
              key.xlab="norm.counts",
              key.ylab="Counts",
              key.par=list(mar=c(3,4,3,1)),
              notecex=3, 
              main=paste(nrow(x_norm),"signif. Diff/ial splicing events\nin ",pair))
    #print(hmp)
    return(ase.filt)
   
}

```


#### PCA of samples on Incl.level 

```{r pca_on_incl_level,echo=T ,eval=F}

mrg_4w_10w <- dplyr::inner_join(ase_mrg_4w,ase_mrg_10w, by="alternative_exon_coordinates")

mrg_4w_10w.sig <- subset(mrg_4w_10w, FDR.x < 0.05 ) #  2088

mrg_4w_10w.incLevels <- cbind(mrg_4w_10w.sig [,grep("IncLevel_4WC_*", colnames(mrg_4w_10w.sig ))],
                              mrg_4w_10w.sig [,grep("IncLevel_4WM_*", colnames(mrg_4w_10w.sig ))],
                              mrg_4w_10w.sig [,grep("IncLevel_10WC_*", colnames(mrg_4w_10w.sig ))],
                              mrg_4w_10w.sig [,grep("IncLevel_10WM_*", colnames(mrg_4w_10w.sig ))] )

mrg_4w_10w.incLevels <- mrg_4w_10w.incLevels %>% drop_na()

colnames(mrg_4w_10w.incLevels) <- gsub("IncLevel_","",colnames(mrg_4w_10w.incLevels))

pca_res <- prcomp(t(mrg_4w_10w.incLevels ), scale=F) 
#do PCA on genes ,not columns/samples
summary(pca_res)
names(pca_res)

var_explained <- pca_res$sdev^2/sum(pca_res$sdev^2)

#png("Plots/PCA_Incl_levels_4W_et_10W.png")
pca_res$x %>%
  as.data.frame %>%
  rownames_to_column("Samples") %>% 
  # create a new column  "Conditions", from samples, without the suffix "r[1-4]
  separate(col=Samples,into = c("Conditions"), sep ="_r[1-4]") %>% 
  ggplot(aes(x=PC1,y=PC2)) + 
  geom_point(aes(color=Conditions),size=3) +
  theme_bw(base_size=10) +
  labs(x=paste0("PC1: ",round(var_explained[1]*100,1),"%"),
       y=paste0("PC2: ",round(var_explained[2]*100,1),"%")) +
  ggtitle("PCA of splicing")
  # + theme(legend.position="top")
#dev.off()



```

##### VolcanoPlot of Incl.Level with dPSI & FDR ==> .x = 4WM/4WC

```{r}

FDR_co = 1.3
dpsi_co = 0.1


condition <- with(mrg_4w_10w, ifelse(dPSI.x > dpsi_co &  
                                       -log10(FDR.x) > FDR_co, "signif_UpSpliced", 
                                       ifelse(dPSI.x < - dpsi_co & 
                                       -log10(FDR.x) > FDR_co, "signif_DownSpliced", 
                                       "Not_Significant")))

colour_values <- c("signif_UpSpliced" = "green4",
                   "signif_DownSpliced"  = "tomato1",
                   "Not_Significant" = "grey15")

png("Plots/Volcano_All_Genes_signif_in_4WM_vs_10WM.png")
vlcpl <- ggplot(data=mrg_4w_10w) + 
          geom_point(aes(x=dPSI.x, y=-log10(FDR.x),colour=condition)) +
          scale_color_manual(name="", values=colour_values) +
          theme(legend.position = "right") +   
          xlab("dPSI") + ylab("-log10(FDR)") + 
          xlim(c(-1.5,1.5)) +  ylim(c(0,20) )  +
          ggtitle("Signif Splicing 4WM-vs-4WC")
print(vlcpl)
dev.off()

```

##### VolcanoPlot of Incl.Level with dPSI & FDR ==> .x = 10WM/10WC

```{r}

FDR_co = 1.3
dpsi_co = 0.1


condition <- with(mrg_4w_10w, ifelse(dPSI.y > dpsi_co &  
                                    -log10(FDR.y) > FDR_co, "signif_UpSpliced", 
                                     ifelse(dPSI.y < - dpsi_co & 
                                    -log10(FDR.y) > FDR_co, "signif_DownSpliced", 
                                       "Not_Significant")))

colour_values <- c("signif_UpSpliced"    = "green4",
                   "signif_DownSpliced"  = "tomato1",
                   "Not_Significant"     = "grey15" )

png("Plots/Volcano_Signif_Splicing_in_10WM-vs-10WC.png")
vlcpl <- ggplot(data=mrg_4w_10w) + 
          geom_point(aes(x=dPSI.y, y=-log10(FDR.y),colour=condition)) +
          scale_color_manual(name="", values=colour_values) +
          theme(legend.position = "right") +   
          xlab("dPSI") + ylab("-log10(FDR)") + 
          xlim(c(-1.5,1.5)) +  ylim(c(0,20) )  +
          ggtitle("Signif Splicing 10WM-vs-10WC")
print(vlcpl)

dev.off()

```


```{r create_piechart_function,echo=F,include=F}

create_pieChart_Splicing <- function(ase, pair, signif.co,psi.co) {
  ##""" Test with : ase = ase.saline; pair="Saline_vs_MBNLdecoy" ; signif.co =0.05 ; psi.co=0.20 """
  
  library(ggplot2)

  ## Count events Signif. splicing : 
  ## count InclDifference(dPSI) = mean(Saline) -  mean(MBNLdecoy)
  ## INCLUDED : PSI < -0.15 
  ## EXCLUDED : PSI > 0.15
  
  signif <- nrow(subset(ase, (FDR < signif.co  &  abs(dPSI) >= psi.co )))
  all_splice <-  nrow(ase)
  
  splice_counts <- data.frame(category=c("All", "Signif & highPSI"),
                              count=c(all_splice,signif))
  
  ## calculate percentage of each count : 
  splice_counts$fraction = round(splice_counts$count / sum(splice_counts$count),2) * 100
  
  # Compute the cumulative percentages (top of each rectangle)
  splice_counts$ymax = cumsum(splice_counts$fraction)
  
  # Compute the bottom of each rectangle
  splice_counts$ymin = c(0, head(splice_counts$ymax, n=-1))
  
  # Compute label position
  # pos <- (splice_counts$ymax + splice_counts$ymin) / 2 # = 49.0 98.5 99.5
  # 98 & 99 are very close, I ll replace them with 97 and 10, left and right of the end of circle.
  splice_counts$labelPosition <- c(50.0, 97.0)
  
  # Compute a good label
  splice_counts$label <- paste0(splice_counts$category,"\n", splice_counts$fraction, " %")
  
  # Make the plot
  dir.create("Plots/",showWarnings = F)
  #png(paste0("Plots/Donut_Plot_Signif_SplicingEvents_",pair,".png"))
    ggplot(splice_counts, aes(ymax=ymax, ymin=ymin, xmax=4, xmin=3, fill=category)) +
      geom_rect() +
      geom_label( x=3.5, aes(y=labelPosition, label=label), size=3) +
      coord_polar(theta="y") + 
      xlim(c(2, 4)) + 
      theme_void()
 #dev.off()
    
}

```


```{r, call_function,echo=F,eval=F}
## call it with ASE table : 
# create_pieChart_Splicing(ase_mrg, pair="Saline_vs_MBNLdecoy",signif.co=0.05, psi.co=0.20)

```


##> Search for common genes  :

```{r,echo=F,eval=F}
ase_mrg_4w.filt$geneSymbol[which(ase_mrg_4w.filt$geneSymbol %in% ase_mrg_10w.filt$geneSymbol)]

ase_mrg_4w.filt$geneSymbol[which(ase_mrg_4w.filt$geneSymbol %in% ase_mrg_10vs4.filt$geneSymbol)]



```

Search for unique /different of one from others: 

```{r ,echo=F,eval=F}
ase_mrg_4w.filt[which(! ase_mrg_4w.filt$geneSymbol %in% ase_mrg_10w.filt$geneSymbol),]

ase_mrg_4w.filt[which(! ase_mrg_4w.filt$geneSymbol %in% ase_mrg_10vs4.filt$geneSymbol),]


```


